---
title: "Adjusted survival curves"
site: distill::distill_website
bibliography: adjcurves.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Background

Unadjusted Kaplan-Meier curves from observational data might be biased because of confounding (@COLE200445).

# Analysis steps of Cole and Hernan 2004

Thanks to Marcel Zwahlen for data preparation.

```{r, echo=F}
main.path <- "/Users/am20q919/Clinical studies/Lectures/JournalClub"
fig.path <- "/Users/am20q919/Clinical studies/Lectures/JournalClub/05_figures/"
```

```{r, echo=F, message=F, warning=F}
library(tidyverse)
library(haven)
library(Hmisc)
library(survey)
library(survival)
library(kableExtra)
library(gtsummary)

data <- read_dta(paste0(paste0(main.path, "/04_prepared_data/"), "11433ColeHernanTableExample.dta"))
```

## Head of data

```{r, echo=F, message=F, warning=F, results='markup'}
head(data)
```

## Descriptive table

```{r, echo=F, message=F, warning=F, results='asis'}
data %>% select(ldh, tx) %>% tbl_summary(by=tx, label=list(tx ~ "Treatment")) %>% as_kable_extra(format = "latex") %>% kableExtra::kable_styling(position = "left", latex_options = c("striped", "repeat_header"))
```

## Kaplan-Meier curve

```{r, echo=T, message=F, warning=F}
mod <- survfit(Surv(time, event) ~ factor(tx), data = data)
coxph(Surv(time, event) ~ factor(tx), data = data)

library(survminer)
plot(mod, col=1:2, data=data)
ggsurvplot(mod, data=data)
?ggsurvplot
```

### IPW modelling

```{r, echo=T, message=F, warning=F, results='markup'}
# IPW denominator
mod <- glm(tx ~ ldh, data=data, family=binomial())
summary(mod)

# OR
data.frame(or=exp(mod$coefficients[2]), lci=exp(confint.default(mod)[2,1]), 
           uci=exp(confint.default(mod)[2,2]))

data$ipw <- NA
# Probabilty of treatment
data$ipw <- predict(mod, data=data, type="response")
# Probabilty of non-treatment
data$ipw[data$tx==0] <- 1-predict(mod, data=data, type="response")[data$tx==0]

# Unstabilized weights
data$ipw_tmp <- 1/data$ipw

library(survey)
svy_design <- svydesign(id=~1, weights=~ipw_tmp, data=data)

table(tx=data$tx, ldh=data$ldh)
svytable(~tx+ldh, design=svy_design)

# Stabilized weights
mod0 <- glm(tx ~ 1, data=data, family=binomial())
data$ipw0 <- predict(mod0, data=data, type="response")
data$ipw0[data$tx==0] <- 1-predict(mod0, data=data, type="response")[data$tx==0]
data$ipw <- data$ipw0/data$ipw
```

### Statement 3

p47: Table 2, column "sw"

```{r, echo=T, message=F, warning=F, results='markup'}
describe(data$ipw)
```

### Statement 3

p47: Table 2, column "Pseudo N"

```{r, echo=T, message=F, warning=F, results='markup'}
svy_design <- svydesign(id=~1, weights=~ipw, data=data)

table(tx=data$tx, ldh=data$ldh)
svytable(~tx+ldh, design=svy_design)
```


### Statement 3

p47: In the pseudopopulation, the indicator of elevated LDH is no longer associated with treatment (odds ratio = 1.0; robust 95% confidence interval: 0.4, 2.7).

```{r, echo=T, message=F, warning=F, results='markup'}
# Robust CIs
mod <- svyglm(tx ~ ldh, data=data, family=binomial(), design=svy_design)
summary(mod)

# OR
data.frame(or=exp(mod$coefficients[2]), lci=exp(confint.default(mod)[2,1]), 
           uci=exp(confint.default(mod)[2,2]))
```


### Statement 3

p47: The estimated IPW hazard ratio comparing treatment S4 against S1—S3 was 1.1 (robust 95% confidence interval: 0.6, 2.1; robust P = 0.78), indistinguishable from the results attained from standard covariate adjustment (adjusted hazard ratio = 1.1; 95% confidence interval: 0.6, 2.1).

```{r, echo=T, message=F, warning=F, results='markup'}
# Adjusted Cox
mod <- coxph(Surv(time, event) ~ factor(tx)+factor(ldh), data=data)
mod
# OR
data.frame(or=exp(mod$coefficients), lci=exp(confint.default(mod)[,1]), 
           uci=exp(confint.default(mod)[,2]))
# IPW adjusted Cox
mod <- svycoxph(Surv(time, event) ~ factor(tx),  design=svy_design)
mod
# OR
data.frame(or=exp(mod$coefficients), lci=exp(confint.default(mod)[,1]), 
           uci=exp(confint.default(mod)[,2]))

# Weighted Kaplan-Meier curves
mod_wght <- svykm(Surv(time, event) ~ factor(tx), design=svy_design)
mod_wght[[1]][2]
plot(mod_wght[[1]], col="black")
lines(mod_wght[[2]], col="red")

library(rsample)
boot_data <- bootstraps(data=data, times=1000, strata=event, apparent=T)

km_split <- function(split) {
  
  mod <- glm(tx ~ ldh, data=analysis(split), family=binomial())
  
  ipw <- predict(mod, data=analysis(split), type="response")
  ipw[analysis(split)$tx==0] <- 1-predict(mod, data=analysis(split), type="response")[analysis(split)$tx==0]

  svy_design <- svydesign(id=~1, weights=~ipw, data=analysis(split))

   mod <- svykm(Surv(time, event) ~ factor(tx), design=svy_design)
  data.frame(group=c(rep(1, length(mod[[1]]$time)), rep(2, length(mod[[2]]$time))), time=c(mod[[1]]$time, mod[[2]]$time), surv=c(mod[[1]]$surv, mod[[2]]$surv))
}

# boot_res <- boot_data %>% mutate(model=map(splits, km_split))
# boot_id0 <- map(boot_res$id, function(x) x) %>% unlist()
# boot_n <- map(boot_res$model, function(x) nrow(x)) %>% unlist()
# boot_coef_group <- map(boot_res$model, function(x) x[,1]) %>% unlist()
# boot_coef_time <- map(boot_res$model, function(x) x[,2]) %>% unlist()
# boot_coef_surv <- map(boot_res$model, function(x) x[,3]) %>% unlist()
# data_gg <- bind_cols(boot=rep(boot_id0, boot_n), group=boot_coef_group, time=boot_coef_time, surv=boot_coef_surv)

# data_gg_agg <- data_gg %>% filter(boot!="Apparent") %>% group_by(group, time) %>% summarise(surv_mean=median(surv), upper=quantile(surv, prob=0.975), lower=quantile(surv, prob=0.025))


# ggplot(data_gg_agg, aes(x=time, y=lower, colour=factor(group)))+geom_point()+geom_point(aes(x=time, y=upper, colour=factor(group)))
```