---
title: "Comparison of hierachical priors in INLA and STAN"
bibliography: hierachicalpriors.bib
site: distill::distill_website
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r base, echo=F, warning=F, message=F}
library(tidyverse)
library(INLA)
library(rstan)
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)
library(Hmisc)

# Main path
pp <-"/Users/am20q919/Clinical studies/Github/Vignettes/HierachicalPriors"
```


# Background

We compare results using different hierarchical priors from INLA and STAN. Specifically, we use default priors and priors suggested by Andrew Gelman (@10.1214/06-BA117A).

# Hierarchical data structure

We assume a hierarchical model
\[
logit(y_{ij})=\alpha+\beta^\top\mathbf{X}+b_i, \quad j\leq J, i\leq n 
\]
with $b_i \sim N(0, \sigma^2)$.

# Priors

Jeffrey's prior for $\sigma^k$ is $f(\sigma^k) \propto 1/\sigma^k$, $k>0$. For, e.g., $k=2$ this is an improper prior distribution because $\int_0^\infty x^{-2}dx=-x^{-2}|_0^\infty=\infty$.

By change-of-variables we get that $\theta=log(\sigma)$

\[
f_{\theta}(\theta)=f(\sigma)\{h^{-1}(\theta)\} \cdot \vert\frac{dh^{-1}(\theta)}{d\theta} \vert=(1/exp(\sigma))\cdot exp(\sigma)=1,
\]

i.e setting a $f(\sigma) \propto 1/\sigma$ is equivalent to $f(log(\sigma)) \propto 1$.

## Aggregated binomial data

```{r data0, echo=T, warning=F, message=F, results="markup"}
library(lme4)
head(cbpp)
```

### Frequentist glmer model

Period as categorical variable

```{r data, echo=T, warning=F, message=F, results="markup"}
gm1 <- glmer(cbind(incidence, size - incidence) ~ factor(period) + (1 | herd), 
             data = cbpp, family = binomial)
summary(gm1)
```

### STAN

STAN model with prior $p(\sigma)=Halfcauchy(0, 25)$.

```{r stan prior1, echo=T, warning=F, message=F}
# library(brms)
# 
# b12.2 <- 
#   brm(data = cbpp, family = binomial,
#       incidence | trials(size) ~ 1 + (1 | herd),
#       prior = c(prior(normal(0, 1000), class = Intercept),
#                 prior(cauchy(0, 25), class = sd)),
#       iter = 4000, warmup = 1000, chains = 4, cores = 4,
#       seed = 12)

### Get contrasts for period
period_factors <- model.matrix(incidence ~ factor(period), data=cbpp)

stan_data <- list(N = nrow(cbpp), G=length(unique(cbpp$herd)), y = cbpp$incidence, 
                  N_i=cbpp$size, g=as.numeric(cbpp$herd), K=ncol(period_factors[,-1]), X=period_factors[,-1])


write("
data {
  int N; // number of obs
  int G; // number of groups
  int K; // number of predictors

  int y[N];
  int N_i[N];
  int g[N]; // map obs to groups
  row_vector[K] X[N];
}

parameters {
  real a_g[G];
  real intercept; // improper uniform prior (-inf, inf) 
  vector[K] beta; // improper uniform prior (-inf, inf) 
  real <lower=0> sigma;
}

model {
    vector[N] p;
    for (i in 1:N) {
        p[i] = inv_logit(intercept+X[i]*beta+a_g[g[i]]);
    }
    y ~ binomial(N_i, p);
    a_g ~ normal(0, sigma);
    intercept ~ normal(0, 1000); // same as INLA
    sigma ~ cauchy(0, 25);
}
"
, file=paste0(pp, "/04_prepared_data/", "StanPrior1.stan"))


stanfile <- paste0(pp, "/04_prepared_data/", "StanPrior1.stan")

fit <- stan(file = stanfile, data = stan_data, warmup = 500, iter = 5000, chains = 4, 
            cores = 6, thin = 1, seed=14749823)


# Get posteriors
posteriors <- as.data.frame(extract(fit))

plot_data <- data.frame(x=density(posteriors$sigma)$x, y=density(posteriors$sigma)$y, type="posterior")
x_prior <- seq(0,max(posteriors$sigma), 0.01)
cauchy_scale <- 25
plot_data <- bind_rows(plot_data, data.frame(x=x_prior, y=dcauchy(x_prior, 0, cauchy_scale), type="prior"))

# Plot posterior distribution of SD of hyperparameter
ggplot(plot_data, aes(x=x, y=y, colour=type, linetype=type))+geom_line()+theme_bw()+scale_color_brewer("", palette="Dark2")+guides(linetype="none")+ylab("Density")+xlab("Hyperparameter sigma")

# 88% credible intervals
alpha <- 0.12
quantile(posteriors$sigma, probs=c(alpha/2, 1-alpha/2))
# P(SD<=q)=0.12
quantile(posteriors$sigma, probs=alpha)

# Probability that hyperparameter!=0
sum(posteriors$sigma>0)/length(posteriors$sigma)
```


```{r stan prior2, echo=T, warning=F, message=F}
write("
data {
  int N; // number of obs
  int G; // number of groups
  int K; // number of predictors

  int y[N];
  int N_i[N];
  int g[N]; // map obs to groups
  row_vector[K] X[N];
}

parameters {
  real a_g[G];
  real intercept; // improper uniform prior (-inf, inf) 
  vector[K] beta; // improper uniform prior (-inf, inf) 
  real <lower=0> sigma; // improper uniform prior (-inf, inf) 
}

model {
    vector[N] p;
    for (i in 1:N) {
        p[i] = inv_logit(intercept+X[i]*beta+a_g[g[i]]);
    }
    y ~ binomial(N_i, p);
    a_g ~ normal(0, sigma);
    intercept ~ normal(0, 1000); // same as INLA
}
"
, file=paste0(pp, "/04_prepared_data/", "StanPrior2.stan"))


stanfile <- paste0(pp, "/04_prepared_data/", "StanPrior2.stan")

fit2 <- stan(file = stanfile, data = stan_data, warmup = 500, iter = 5000, chains = 4, 
            cores = 6, thin = 1, seed=14749823)


# Get posteriors
posteriors2 <- as.data.frame(extract(fit2))
```


```{r stan prior3, echo=T, warning=F, message=F}
write("
data {
  int N; // number of obs
  int G; // number of groups
  int K; // number of predictors

  int y[N];
  int N_i[N];
  int g[N]; // map obs to groups
  row_vector[K] X[N];
}

parameters {
  real a_g[G];
  real intercept; // improper uniform prior (-inf, inf) 
  vector[K] beta; // improper uniform prior (-inf, inf) 
  real <lower=0> sigma;
}

model {
    vector[N] p;
    for (i in 1:N) {
        p[i] = inv_logit(intercept+X[i]*beta+a_g[g[i]]);
    }
    y ~ binomial(N_i, p);
    a_g ~ normal(0, sigma);
    intercept ~ normal(0, 1000); // same as INLA
    sigma ~ inv_gamma(1, 0.00005);
}
"
, file=paste0(pp, "/04_prepared_data/", "StanPrior3.stan"))


stanfile <- paste0(pp, "/04_prepared_data/", "StanPrior3.stan")

fit3 <- stan(file = stanfile, data = stan_data, warmup = 500, iter = 5000, chains = 4, 
            cores = 6, thin = 1, seed=14749823)


# Get posteriors
posteriors3 <- as.data.frame(extract(fit3))
```

```{r stan prior4, echo=T, warning=F, message=F}
write("
data {
  int N; // number of obs
  int G; // number of groups
  int K; // number of predictors

  int y[N];
  int N_i[N];
  int g[N]; // map obs to groups
  row_vector[K] X[N];
}

parameters {
  real a_g[G];
  real intercept; // improper uniform prior (-inf, inf) 
  vector[K] beta; // improper uniform prior (-inf, inf) 
  real <lower=0> sigma;
}

model {
    vector[N] p;
    for (i in 1:N) {
        p[i] = inv_logit(intercept+X[i]*beta+a_g[g[i]]);
    }
    y ~ binomial(N_i, p);
    a_g ~ normal(0, sigma);
    intercept ~ normal(0, 1000); // same as INLA
    sigma ~ cauchy(0, 0.1);
}
"
, file=paste0(pp, "/04_prepared_data/", "StanPrior4.stan"))


stanfile <- paste0(pp, "/04_prepared_data/", "StanPrior4.stan")

fit4 <- stan(file = stanfile, data = stan_data, warmup = 500, iter = 5000, chains = 4, 
            cores = 6, thin = 1, seed=14749823)


# Get posteriors
posteriors4 <- as.data.frame(extract(fit4))
```


### INLA

Note: INLA is internally parametrised on the log precision. When $log(\tau)$ is logGamma(a,b) distributed then (i.e. the log precision) then $\tau$ is Gamma(a,b) distributed with density
\[
p(\tau)=\frac{b^a}{\Gamma(a)}\tau^{a-1}exp(-b\tau),
\]
where $b$ is the inverse scale parameter (see https://inla.r-inla-download.org/r-inla.org/doc/prior/prior-loggamma.pdf). When $\tau \sim Gamma(a,s)$ then $\tau^{-1} \sim InvGamma(a,s)$.

Note: The base R functions (dgamma, pgamma, qgamma and rgamma) are parametrised as
\[
p(\tau)=\frac{1}{b^a\Gamma(a)}\tau^{a-1}exp(-s\tau),
\]
where $s=1/b$ is the rate parameter.

```{r inla, echo=T, warning=F, message=F}
library(INLA)
library(brinla)

HC.prior  = "expression:
  sigma = exp(-theta/2);
  gamma = 25;
  log_dens = log(2) - log(pi) - log(gamma);
  log_dens = log_dens - log(1 + (sigma / gamma)^2);
  log_dens = log_dens - log(2) - theta / 2;
  return(log_dens);
"

HC.prior2  = "expression:
  sigma = exp(-theta/2);
  gamma = 0.1;
  log_dens = log(2) - log(pi) - log(gamma);
  log_dens = log_dens - log(1 + (sigma / gamma)^2);
  log_dens = log_dens - log(2) - theta / 2;
  return(log_dens);
"

# Note: UN.prior is unrestricted!
UN.prior = "expression:
  log_dens = 0 - log(2) - theta / 2;
  return(log_dens);
"

prior.list = list(
  prior1 = list(prec = list(prior = HC.prior)),
  prior2 = list(prec = list(prior = UN.prior)),
  default = list(prec = list(prior = "loggamma", param = c(1, 0.00005))),
  prior3 = list(prec = list(prior = HC.prior2))

) 

intercept_fixed_prio <- list(mean.intercept=0, prec.intercept=1/1000)

inla_models <- lapply(prior.list, function(tau.prior) {
  inla(incidence ~ factor(period)+f(herd, model="iid", hyper=tau.prior), data=cbpp, family="binomial", control.predictor = list(link=1, compute = TRUE), control.compute = list(dic=T, cpo=T, waic=T), control.fixed=intercept_fixed_prio, Ntrials=size)
})
```

```{r inla res, echo=T, warning=F, message=F}
results <- c()
results <- bind_rows(results, data.frame(bri.hyper.sd(inla_models$prior1$marginals.hyperpar[[1]]), type="INLA: HC(0, 25)"))
results <- bind_rows(results, data.frame(bri.hyper.sd(inla_models$prior2$marginals.hyperpar[[1]]), type="INLA: Improper uniform"))
results <- bind_rows(results, data.frame(bri.hyper.sd(inla_models$default$marginals.hyperpar[[1]]), type="INLA: Gamma(1, 0.00005)"))
results <- bind_rows(results, data.frame(bri.hyper.sd(inla_models$prior3$marginals.hyperpar[[1]]), type="INLA: HC(0, 0.1)"))
results <- bind_rows(results, data.frame(x=density(posteriors$sigma)$x, y=density(posteriors$sigma)$y, type="STAN: HC(0, 25)"))
results <- bind_rows(results, data.frame(x=density(posteriors2$sigma)$x, y=density(posteriors2$sigma)$y, type="STAN: Improper uniform"))
results <- bind_rows(results, data.frame(x=density(posteriors3$sigma)$x, y=density(posteriors3$sigma)$y, type="STAN: Gamma(1, 0.00005)"))
results <- bind_rows(results, data.frame(x=density(posteriors4$sigma)$x, y=density(posteriors4$sigma)$y, type="STAN: HC(0, 0.1)"))

ggplot(results, aes(x=x, y=y, colour=type))+geom_line()+coord_cartesian(ylim=c(0,3))+facet_wrap(~type)+geom_vline(xintercept=sqrt(summary(gm1)$varcor[[1]]), linetype="dashed")+labs(caption="Dashed vertical line: Estimated SD of hyperparameter from GLMER")+theme_bw()+theme(legend.position = "bottom")+scale_color_brewer("", palette="Dark2")
```