---
title: "Immortal time bias"
author:
  - name: André Moser, CTU Bern, University of Bern
    orcid_id: 0000-0001-7178-6539
output: 
  distill::distill_article:
    toc: true
    number_sections: true    
    toc_float: true
bibliography: immortal.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Review

This vignette has not been (yet) reviewed by other statisticians `r emo::ji("face screaming in fear")`. `r emo::ji("skull")` You read it on your own risk `r emo::ji("skull")`.

# Background

Immortal time bias might arise, if - for example - individuals are assigned to an exposure or treatment group based on covariate information after time zero (@hernan). Consequently, group assignment is not aligned with time zero and individuals inherently experience no outcome event for a specific time period (they are immortal when the outcome of interest is time to death). Immortal time bias is common when analyzing observational studies (@suissa, @levesque). Other terminology for immortal time bias is *time dependent bias* or *survival bias*. In this vignette we introduce several methods to address immortal time bias discussed in the articles by @zhou and @robins.

# Example

We consider a study population of 10,000 patients admitted to an emergency unit. Patients with a more severe admission status are treated more quickly. Clinicians are interested whether the time from admission to time to treatment affect survival and group patients into two treatment groups: Short and long time to treatment after emergency admission. Group assignment is based on the median time to treatment.

- Research question: **What is the effect of the time to treatment group assignment on survival?**
- Study design: Cohort study
- Outcome of interest: Time to death or end of follow-up
- Observation time: Admission to end of follow-up
- Predictor of interest: Time to treatment after admission
- Confounders: Severity

The used variables from the data are:

| Variable | Definition | Coding |
| --- | --- | --- |
| group | Group assignment | 1=Long (exposed), 0=Short (unexposed) |
| severity | Severity status | 1=Severe, 0=Non-severe |
| death | Death | 1=Death, 0=Alive |
| time_to_treatment | Time to treatment (days) | Non-negative number |
| fup | Follow-up time (days) | Non-negative number |

## Knowledge from the crystal ball

Because we simulated the data, we know that group assignment has no effect on mortality.

# Analysis strategy

@zhou, @karim, @maringe, @robins and @hernan discuss different analysis strategies for addressing (or not addressing) immortal time bias. In this vignette we summarize the investigated methods from the above mentioned articles:

| Method | Definition |
| --- | --- |
| 1 | Group assignment at time zero |
| 2 | Random time to treatment assignment |
| 3 | Matched time to treatment assignment |
| 4 | Time zero at end of treatment |
| 5 | Time dependent treatment |
| 6 | Cloning |

```{r, echo=T, message=F, warning=F}
# Required packages
library(tidyverse)
library(survey)
library(survival)
library(survminer)
```

```{r, echo=F, message=F, warning=F}
library(gtsummary)
```

```{r, echo=F, message=F, warning=F, results='asis'}
n <- 10000

set.seed(1)

# Time to treatment: Mean 10 days
time_to_treatment <- rweibull(n, 5, 10)
# Two equal groups based on time to treatment
quantile_info <- median(time_to_treatment)
group <- ifelse(time_to_treatment>median(time_to_treatment), 1, 0)
# More severe patients are treated more quickly
severity <- ifelse(runif(n)<plogis(qlogis(0.8)+log(0.2)*(group==1)), 1, 0)
# Hazard of dying: Independent of time to treatment and severity
haz <- 0.05
# Time to death: Exponential with rate=haz
time_to_death <- -log(runif(n))/haz
# Censoring: Mean 20 days
censored <- rweibull(n, 1, 20)
# Follow-up time: Patients do not die until they are treated
fup <- pmin(time_to_treatment+time_to_death, time_to_treatment+censored)
# Death indicator: Patients do not die until they are treated
death <- ifelse(time_to_treatment+time_to_death<=time_to_treatment+censored, 1, 0)
# Data
data <- data.frame(id=1:n, fup, time_to_treatment, censored, time_to_death, death, group, severity)
```

## Descriptive table

The table below shows a descriptive summary of the study population, by group assignment.

```{r, echo=F, message=F, warning=F, results='asis'}
data$group <- factor(data$group, levels=0:1, labels=c("Short", "Long"))
```

```{r, echo=F, message=F, warning=F, results='asis'}
data %>% select(!c(id, censored, time_to_death)) %>% 
  tbl_summary(by=group, label=list(group ~ "Group assignment"))
```

```{r, echo=F, message=F, warning=F, results='asis'}
data$group <- as.numeric(data$group)-1
```

```{r, echo=F, message=F, warning=F, results='asis'}
data$group_lab <- NULL
```

## Method 1: Group assignment at time zero

Time zero was time of emergency admission. Treatment was assigned according to whether 
patients had a short or long time to treatment. No patients were excluded from the analysis.

```{r, echo=T, message=F, warning=F}
# Kaplan-Meier survival curves
mod <- survfit(Surv(fup, death)~factor(group), data=data)
ggsurvplot(mod, data=data, palette=c("#CC0000", "black"), censor=F, pval=T, xlab="Days from emergency admission")
```

## Method 2: Random time to treatment assignment

Treatment was assigned according to whether patients had a short or long time to treatment. 
Time zero was set at start of treatment, but time to treatment for the short group was 
randomly replaced by a time to treatment from the time to treatment range of the long group. 

```{r, echo=T, message=F, warning=F}
data_method2 <- data

# New time to treament for short group
data_method2$time_to_treatment_new <- data_method2$time_to_treatment
data_method2$time_to_treatment_new[data_method2$group==0] <- 
  runif(sum(data_method2$group==0), median(time_to_treatment), max(time_to_treatment))

# Exclusion of patients:
# If original follow-up time of patients < random treatment time
data_method2$exclude <- 0
# Patients in the short group who died
id_short <- data_method2$group==0 & data_method2$death==1
data_method2$exclude[data_method2$group==0 & data_method2$death==1] <- 
  ifelse(data_method2$fup[id_short]<data_method2$time_to_treatment_new[id_short], 1, 0)
```

`r sum(data_method2$exclude)` (`r paste0(round(sum(data_method2$exclude)/nrow(data_method2)*100, 1), "%")`) individuals from the short group who died before time zero were excluded.

```{r, echo=T, message=F, warning=F}
data_method2 <- data_method2 %>% filter(exclude==0)

# Time zero: Start of treatment
# Difference between new random treatment time - original treatment time (=zero for long group)
data_method2$fup <- pmin(data_method2$time_to_treatment_new-data_method2$time_to_treatment+data_method2$time_to_death, 
                         data_method2$time_to_treatment_new-data_method2$time_to_treatment+data_method2$censored)
# Replace death indicator
data_method2$death <- ifelse(data_method2$time_to_treatment_new-data_method2$time_to_treatment+data_method2$time_to_death<=
                               data_method2$time_to_treatment_new-data_method2$time_to_treatment+data_method2$censored, 1, 0)

# Kaplan-Meier survival curves
mod <- survfit(Surv(fup, death)~factor(group), data=data_method2)
ggsurvplot(mod, data=data_method2, palette=c("#CC0000", "black"), censor=F, pval=T, xlab="Days from start of treatment")
```

## Method 3: Matched time to treatment assignment

Treatment was assigned according to whether patients had a short or long time to treatment. Time zero was set at start of treatment, but time to treatment for the short group was randomly replaced by a time to treatment from the time to treatment range of the long group. This methods corrects for time to treatment imbalances between the two groups (in contrast to method 2).

```{r, echo=T, message=F, warning=F}
data_method3 <- data
# Get time to treatment from long group
sample_time <- data_method3$time_to_treatment[data_method3$group==1]
# New time to treatment for short group: Matching from long group
data_method3$time_to_treatment_new <- data_method3$time_to_treatment
data_method3$time_to_treatment_new[data_method3$group==0] <- 
  sample(sample_time, length(data_method3$time_to_treatment[data_method3$group==0]))

# Exclusion of patients
data_method3$exclude <- 0
data_method3$exclude[data_method3$group==0 & data_method3$death==1] <- 
  ifelse(data_method3$fup[data_method3$group==0 & data_method3$death==1]<
           data_method3$time_to_treatment_new[data_method3$group==0 & data_method3$death==1], 1, 0)
```

`r sum(data_method3$exclude)` (`r paste0(round(sum(data_method3$exclude)/nrow(data_method3)*100, 1), "%")`) individuals from the short group who die before time zero were excluded.

```{r, echo=T, message=F, warning=F}
data_method3 <- data_method3 %>% filter(exclude==0)

# Time zero: Start of treatment
# Difference between new matched treatment time - original treatment time (=zero for long group)
data_method3$fup <- pmin(data_method3$time_to_treatment_new-data_method3$time_to_treatment+data_method3$time_to_death, 
                         data_method3$time_to_treatment_new-data_method3$time_to_treatment+data_method3$censored)
data_method3$fup <- pmin(data_method3$time_to_death, data_method3$censored)
# Replace death indicator
data_method3$death <- ifelse(data_method3$time_to_treatment_new-data_method3$time_to_treatment+data_method3$time_to_death<=
                               data_method3$time_to_treatment_new-data_method3$time_to_treatment+data_method3$censored, 1, 0)

# Kaplan-Meier survival curves
mod <- survfit(Surv(fup, death)~factor(group), data=data_method3)
ggsurvplot(mod, data=data_method3, palette=c("#CC0000", "black"), censor=F, 
           pval=T, xlab="Days from start of treatment")
```

## Method 4: Time zero at end of treatment

Treatment was assigned according to whether patients had a short or long time to treatment. End of treatment was defined at the last observed time to treatment (`r round(max(time_to_treatment), 1)` days). Individuals were followed-up from the end of treatment duration (time zero).

```{r, echo=T, message=F, warning=F}
data_method4 <- data
data_method4$exclude <- ifelse(data_method4$fup<max(time_to_treatment), 1, 0)
```

`r sum(data_method4$exclude)` (`r paste0(round(sum(data_method4$exclude)/nrow(data_method4)*100, 1), "%")`) individuals from both groups with a follow-uo time smaller than end of treatment duration were excluded.

```{r, echo=T, message=F, warning=F}
data_method4 <- data_method4 %>% filter(exclude==0)
# Shift time to follow-up
data_method4$fup <- data_method4$fup-max(time_to_treatment)

# Kaplan-Meier survival curves
mod <- survfit(Surv(fup, death)~factor(group), data=data_method4)
ggsurvplot(mod, data=data_method4, palette=c("#CC0000", "black"), censor=F, 
           pval=T, xlab="Days from end of treatment duration")
```

## Method 5: Time-dependent treatment

Treatment assignment was 0 ("short") as long as a patient had a time to treatment
smaller than the median time to treatment, and 1 ("long") otherwise. No individuals were excluded.

```{r, echo=T, message=F, warning=F}
library(splitstackshape)
# For "smoother" curves: Fup*10
multiplier <- 10
data$fup2 <- data$fup*multiplier
data_discrete_surv <- expandRows(data, count="fup2", drop=F) %>% arrange(id)
# Count indicator: How many follow-up days per individuals
data_discrete_surv <- data_discrete_surv %>% group_by(id) %>% mutate(ind=1, time=cumsum(ind)-1)
data_discrete_surv$ind <- NULL
# Maximal follow-up day
data_discrete_surv <- data_discrete_surv %>% group_by(id) %>% mutate(max_time=max(time))
# Correct death
data_discrete_surv$death[data_discrete_surv$time < data_discrete_surv$max_time] <- 0
# Shift days
data_discrete_surv <- data_discrete_surv %>% group_by(id) %>% mutate(time2=lead(time))
# Fill last day
data_discrete_surv <- data_discrete_surv %>% group_by(id) %>% fill(time2)
data_discrete_surv$time2[data_discrete_surv$time==data_discrete_surv$max_time] <- 
  data_discrete_surv$time2[data_discrete_surv$time==data_discrete_surv$max_time]+1

# Create time-dependent treatment
data_discrete_surv <- data_discrete_surv %>% group_by(id) %>% mutate(group_timedependent=0)
data_discrete_surv$group_timedependent[data_discrete_surv$time>=
            data_discrete_surv$time_to_treatment*multiplier & data_discrete_surv$group==1] <- 1
# Kaplan-Meier survival curves
mod_km <- survfit(Surv(time=time/multiplier, time2=time2/multiplier, event=death)~
                    group_timedependent, data=data_discrete_surv, cluster = id)
ggsurvplot(mod_km, data=data_discrete_surv, palette=c("#CC0000", "black"), 
           censor=F, pval=F, xlab="Days from emergency admission")

# Cox model
cox_fit <- coxph(Surv(time=time/multiplier, time2=time2/multiplier, event=death)~
                   group_timedependent, data=data_discrete_surv, cluster = id)
```

The p-value from a Cox proportional hazard model comparing the two survival curves was p=`r round(summary(cox_fit)$coef[, "Pr(>|z|)"], 3)` (p-value from Schoenfeld test p=`r round(cox.zph(cox_fit)$table[1,3], 3)`).

## Method 6: Cloning

Each patient received the actual observed exposure or treatment and was cloned, receiving the not observed exposure or treatment. This led to two pseudopopulations, each with a size of `r n`` patients. Patients in one pseudopopulation (those having a short time to treatment duration) were censored if the time to treatment is longer than the median time to treatment. Patients in the pseudopopulation with a long time to treatment duration were censored if patients were alive and had a time to treatment shorter than the median time to treatment. Censoring can be seen as a protocol deviation (@maringe). Artificial censoring is addressed by inverse probability weighting (@robins, @hernan). No patients were excluded from the analysis.

```{r, echo=F, message=F, warning=F}
# Short group clone
data_clone0 <- data
# Long group clone
data_clone1 <- data

# Short group clone
# Censor if follow-up time > median time to treatment
data_clone0$censored <- 0
data_clone0$censored[data_clone0$group==1] <- 1
data_clone0$death[data_clone0$group==1] <- 0
data_clone0$fup[data_clone0$group==1] <- median(time_to_treatment)
data_clone0$group <- 0

# Long group clone
# Censor if follow-up time < median time to treatment and alive
data_clone1$censored <- 0
data_clone1$censored[data_clone0$group==0 & data_clone0$death==0] <- 1
data_clone1$group <- 1

# Combine both cloning data sets
data_clone <- bind_rows(data_clone0, data_clone1)

# Construction of inverse probability weights (IPW) for censoring
mod <- glm(censored ~ severity, data=data_clone, family=binomial())

data_clone$ipw <- NA
# Probability of being censored
data_clone$ipw <- predict(mod, data=data_clone, type="response")
# Probability of being censored
data_clone$ipw[data_clone$severity==0] <- 1-predict(mod, data=data_clone, type="response")[data_clone$severity==0]

# Set survey design
svy_design <- svydesign(id=~1, weights=~ipw, data=data_clone)

# IPW adjusted Kaplan-Meier
km_fit <- svykm(Surv(fup, death) ~ group, design=svy_design)

km_df <- data.frame(time=km_fit$`1`$time, surv=km_fit$`1`$surv, strata="group=1")
km_df <- bind_rows(km_df, data.frame(time=km_fit$`0`$time, surv=km_fit$`0`$surv, strata="group=0"))
ggsurvplot_df(km_df, palette=c("#CC0000", "black"), censor=F, pval=F, xlab="Days from emergency admission")

# Cox model
cox_fit <- svycoxph(Surv(fup, death) ~ group, design=svy_design)
```

The p-value from a Cox proportional hazard model comparing the two survival curves was p=`r round(summary(cox_fit)$coef[, "Pr(>|z|)"], 3)` (p-value from Schoenfeld test p=`r round(cox.zph(cox_fit)$table[1,3], 3)`).

# Conclusion

Immortal time bias is common in time to event analysis of observational data. In the present vignette we presented several methods for addressing immortal time bias. Similar to @zhou we conclude that the investigated methods 1 and 2 did not adequately address immortal time bias. In contrast to @zhou we found that method 3 did not adequately address immortal time bias too, which was also highlighted in @karim. Methods 4 to 6 were able to address immortal time bias. We recommend to compare some of the suggested methods in a real world analysis in sensitivity analyses and to be aware of their limitations (for example, exclusion of patients which leads to selection bias).

# Data simulation

```{r, echo=T, message=F, warning=F, results='asis'}
n <- 10000

set.seed(1)

# Time to treatment: Mean 10 days
time_to_treatment <- rweibull(n, 5, 10)
# Two equal groups based on time to treatment
quantile_info <- median(time_to_treatment)
group <- ifelse(time_to_treatment>median(time_to_treatment), 1, 0)
# More severe patients are treated more quickly
severity <- ifelse(runif(n)<plogis(qlogis(0.8)+log(0.2)*(group==1)), 1, 0)
# Hazard of dying: Independent of time to treatment and severity
haz <- 0.05
# Time to death: Exponential with rate=haz
time_to_death <- -log(runif(n))/haz
# Censoring: Mean 20 days
censored <- rweibull(n, 1, 20)
# Follow-up time: Patients do not die until they are treated
fup <- pmin(time_to_treatment+time_to_death, time_to_treatment+censored)
# Death indicator: Patients do not die until they are treated
death <- ifelse(time_to_treatment+time_to_death<=time_to_treatment+censored, 1, 0)
# Data
data <- data.frame(id=1:n, fup, time_to_treatment, censored, time_to_death, death, group, severity)
```
