---
title: "Immortal time bias"
author:
  - name: André Moser, CTU Bern, University of Bern
    orcid_id: 0000-0001-7178-6539
output: 
  distill::distill_article:
    toc: true
    number_sections: true    
    toc_float: true
bibliography: immortal.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Background

Immortal time bias might arise, if - for example - individuals are assigned to an exposure or treatment group (and on information) after time zero (@hernan). Consequently, group assignment is not aligned with time zero and individuals inherently experience no outcome event for a specific time period (they are immortal when the outcome of interest is time to death). Immortal time bias is common when analyzing observational studies (@suissa, @levesque). Other terminology for immortal time bias is *time dependent bias* or *survival bias*. In this vignette we introduce several methods to address immortal time bias (@zhou, @robins).

# Example

We consider a study population of 10,000 patients admitted to an emergency unit. Patients with a more severe admission status are treated more quickly. Clinicians are interested whether the time from admission to time to treatment affect survival and group patients into two treatment groups: Short and long time to treatment after emergency admission.

- Research question: **What is the effect of the time to treatment group assignment on survival?**
- Study design: Cohort study
- Outcome of interest: Time to death or end of follow-up
- Observation time: Admission to end of follow-up
- Predictor of interest: Time to treatment after admission
- Confounders: Severity

The used variables from the data are:

| Variable | Definition | Coding |
| --- | --- | --- |
| group | Group assignment | 0=Short, 1=Long |
| severity | Severity status | 1=Severe, 0=Non-severe |
| death | Death | 1=Death, 0=Alive |
| fup | Follow-up time | Non-negative number |

## Knowledge from the crystal ball

Because we simulated the data, we know that group assignment has no effect on mortality.

# Analysis strategy

@zhou, @karim, @robins and @hernan discuss different analysis strategies for addressing (or not adressing) immortal time bias. IN this vignette we summarise the investigated methods of the mentioned authors:


```{r, echo=T, message=F, warning=F}
# Required packages
library(tidyverse)
library(survey)
library(survival)
library(gtsummary)
library(survminer)
```

```{r, echo=F, message=F, warning=F, results='asis'}
n <- 10000

set.seed(1)

# Time to treatment: Mean 10 days
time_to_treatment <- rweibull(n, 5, 10)
# Two equal groups based on time to treatment
quantile_info <- median(time_to_treatment)
group <- ifelse(time_to_treatment>median(time_to_treatment), 1, 0)
# More severe patients are treated more quickly
severity <- ifelse(runif(n)<plogis(qlogis(0.8)+log(0.2)*(group==1)), 1, 0)
# Hazard of dying: Independent of time to treatment and severity
haz <- 0.05
# Time to death: Exponential with rate=haz
time_to_death <- -log(runif(n))/haz
# Censoring: Mean 10 days
censored <- rweibull(n, 1, 10)
# Follow-up time: Patients do not die until they are treated
fup <- pmin(time_to_treatment+time_to_death, time_to_treatment+censored)
# Death indicator: Patients do not die until they are treated
death <- ifelse(time_to_treatment+time_to_death<=time_to_treatment+censored, 1, 0)
# Data
data <- data.frame(id=1:n, fup, time_to_treatment, censored, time_to_death, death, group, severity)
```

## Descriptive table

The table below shows a descriptive summary of the study population, by group assignment.

```{r, echo=F, message=F, warning=F, results='asis'}
data$group_lab <- factor(data$group, levels=0:1, labels=c("Short", "Long"))
```

```{r, echo=T, message=F, warning=F, results='asis'}
data %>% select(!c(id, censored, time_to_death)) %>% 
  tbl_summary(by=group_lab, label=list(group ~ "Group assignment"))
```

```{r, echo=F, message=F, warning=F, results='asis'}
data$group_lab <- NULL
```

# Immortal time bias: Wrong approaches

This section shows two approaches which do not correct for immortal time bias (@zhou).

## Method 1: Fixed group assignment at time zero

Time zero is time of emergency admission. Treatment is assigned according to whether 
patients had a short or long time to treatment.

```{r, echo=T, message=F, warning=F}
mod <- survfit(Surv(fup, death)~factor(group), data=data)
ggsurvplot(mod, data=data, palette=c("#CC0000", "black"), censor=F, pval=T)
```

## Method 2: Randomly assign time zero for short group

Treatment is assigned according to whether patients had a short or long time to treatment. 
Time zero is set at start of treatment, but time to treatment for the short group is 
randomly replaced by a time to treatment from the time to treatment range of the long group. 
Individuals from the short group who die before time zero are excluded.

```{r, echo=T, message=F, warning=F}
data_method2 <- data
# New time to treament for short group
data_method2$time_to_treatment_new <- data_method2$time_to_treatment
data_method2$time_to_treatment_new[data_method2$group==0] <- 
  runif(length(data_method2$time_to_treatment_new[data_method2$group==0]), median(time_to_treatment), max(time_to_treatment))

# Exclusion of patients
data_method2$exclude <- 0
data_method2$exclude[data_method2$group==0 & data_method2$death==1] <- 
  ifelse(data_method2$fup[data_method2$group==0 & data_method2$death==1]<data_method2$time_to_treatment_new[data_method2$group==0 & data_method2$death==1], 1, 0)

data_method2 <- data_method2 %>% filter(exclude==0)

# Time zero: Start of treatment
data_method2$fup <- pmin(data_method2$time_to_treatment_new-data_method2$time_to_treatment+data_method2$time_to_death, data_method2$time_to_treatment_new-data_method2$time_to_treatment+data_method2$censored)
# Replace death indicator
data_method2$death <- ifelse(data_method2$time_to_treatment_new-data_method2$time_to_treatment+data_method2$time_to_death<=data_method2$time_to_treatment_new-data_method2$time_to_treatment+data_method2$censored, 1, 0)

mod <- survfit(Surv(fup, death)~factor(group), data=data_method2)
ggsurvplot(mod, data=data_method2, palette=c("#CC0000", "black"), censor=F, pval=T, xlab="Time from start of treatment")
```

# Immortal time bias: Approaches correcting for bias

This section shows two approaches which might correct for immortal time bias (@zhou, @robins).

## Method 3: Time zero at end of treatment

Treatment is assigned according to whether patients had a short or long time to treatment. Individuals are followed-up from the end of treatment duration.

```{r, echo=T, message=F, warning=F}
data_method3 <- data
data_method3$excluded <- ifelse(data_method3$fup<max(time_to_treatment), 1, 0)
data_method3 <- data_method3 %>% filter(excluded==0)

data_method3$fup <- data_method3$fup-max(time_to_treatment)

mod <- survfit(Surv(fup, death)~factor(group), data=data_method3)
ggsurvplot(mod, data=data_method3, palette=c("#CC0000", "black"), censor=F, pval=T, xlab="Time from end of treatment duration")
```

## Method 4: Matched treatment duration

Treatment is assigned according to whether patients had a short or long time to treatment. Time zero is set at start of treatment, but time to treatment for the short group is randomly replaced by a time to treatment from the time to treatment range of the long group. Individuals from the short group who die before time zero are excluded. This methods corrects for time to treatment imbalances (compared to method 2).

```{r, echo=T, message=F, warning=F}
data_method4 <- data
# Get time to treatment from long group
sample_time <- data_method4$time_to_treatment[data_method4$group==1]
# New time to treament for short group
data_method4$time_to_treatment_new <- data_method4$time_to_treatment
data_method4$time_to_treatment_new[data_method4$group==0] <- sample(sample_time, length(data_method4$time_to_treatment[data_method4$group==0]))

# Exclusion of patients
data_method4$exclude <- 0
data_method4$exclude[data_method4$group==0 & data_method4$death==1] <- ifelse(data_method4$fup[data_method4$group==0 & data_method4$death==1]<data_method4$time_to_treatment_new[data_method4$group==0 & data_method4$death==1], 1, 0)

data_method4 <- data_method4 %>% filter(exclude==0)

# Time zero: Start of treatment
data_method4$fup <- pmin(data_method4$time_to_treatment_new-data_method4$time_to_treatment+data_method4$time_to_death, data_method4$time_to_treatment_new-data_method4$time_to_treatment+data_method4$censored)
data_method4$fup <- pmin(data_method4$time_to_death, data_method4$censored)
# Replace death indicator
data_method4$death <- ifelse(data_method4$time_to_treatment_new-data_method4$time_to_treatment+data_method4$time_to_death<=data_method4$time_to_treatment_new-data_method4$time_to_treatment+data_method4$censored, 1, 0)

mod <- survfit(Surv(fup, death)~factor(group), data=data_method4)
ggsurvplot(mod, data=data_method4, palette=c("#CC0000", "black"), censor=F, pval=T, xlab="Time from start of treatment")
```

## Method 5: Time-dependent treatment

Treatment assignment is 0 ("short") as long as a patient has a time to treatment
smaller than the median time to treatment, and 1 ("long") otherwise.

```{r, echo=T, message=F, warning=F}
library(splitstackshape)
# For "smoother" curves: Fup*10
multiplier <- 10
data$fup2 <- data$fup*multiplier
data_discrete_surv <- expandRows(data, count="fup2", drop=F) %>% arrange(id)
# Count indicator: How many follow-up days per individuals
data_discrete_surv <- data_discrete_surv %>% group_by(id) %>% mutate(ind=1, time=cumsum(ind)-1)
data_discrete_surv$ind <- NULL
# Maximal follow-up day
data_discrete_surv <- data_discrete_surv %>% group_by(id) %>% mutate(max_time=max(time))
# Correct death
data_discrete_surv$death[data_discrete_surv$time < data_discrete_surv$max_time] <- 0
# Shift days
data_discrete_surv <- data_discrete_surv %>% group_by(id) %>% mutate(time2=lead(time))
# Fill last day
data_discrete_surv <- data_discrete_surv %>% group_by(id) %>% fill(time2)
data_discrete_surv$time2[data_discrete_surv$time==data_discrete_surv$max_time] <- data_discrete_surv$time2[data_discrete_surv$time==data_discrete_surv$max_time]+1

# Create time-dependent treatment
data_discrete_surv <- data_discrete_surv %>% group_by(id) %>% mutate(group_timedependent=0)
data_discrete_surv$group_timedependent[data_discrete_surv$time>=data_discrete_surv$time_to_treatment*multiplier & data_discrete_surv$group==1] <- 1

mod_km <- survfit(Surv(time=time/multiplier, time2=time2/multiplier, event=death)~group_timedependent, data=data_discrete_surv, cluster = id)
ggsurvplot(mod_km, data=data_discrete_surv, palette=c("#CC0000", "black"), censor=F, pval=F, xlab="Time from emergency admission")
```

## Method 6: Cloning

```{r, echo=F, message=F, warning=F}
data_clone0 <- data
data_clone1 <- data

data_clone0$censored <- 0
data_clone0$censored[data_clone0$fup>=median(time_to_treatment)] <- 1

data_clone0$death[data_clone0$fup>=median(time_to_treatment)] <- 0
data_clone0$fup[data_clone0$fup>=median(time_to_treatment)] <- median(time_to_treatment)
data_clone0$group <- 0

data_clone1$censored <- 0
data_clone1$censored[data_clone0$fup<median(time_to_treatment)] <- 1
data_clone1$group <- 1

data_clone <- bind_rows(data_clone0, data_clone1)

#mod <- survfit(Surv(fup, death)~factor(group), data=data_clone)
#ggsurvplot(mod, data=data, palette=c("#CC0000", "black"), censor=F, pval=T)
```

# Conclusion

Immortal time bias is common in time to event analysis of observational data. In the present vignette we presented several methods for addressing immortal time bias. Similar to @zhou we conclude that the investigated methods 1 and 2 do not adequately address immortal time bias. In contrast to @zhou we found that method 4 did not adequatly address immortal time bias, which was also highlighted in @karim.

